---
title: "Paper 2 (Possible Gender Differences) Appendix"
author: "Casey Lee"
date: "2025-02-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
```

## Overview  
Paper 2 appendix provides an analysis of the possible gender differences in overconfidence. We will be creating two regression models (E(Y | intel_theory, attn_to) and E(Y | intel_theory, attn_to, gender):  
1. See if the gender variable fits the first model and plot its fitted values.  
2. Create plots to see if there is an unexplained variation potentially explained by gender.  
3. Calculate in-sample loss for the versions of the first and second models to assess whether the relationship of overplacement to intelligence theory and experimental condition differed for men and women.

We are using the merged file combining the holdout and the original sample to reinforce a paper model with and without a contribution from gender. Then we create two regression models: We find the in-sample loss calculations based on cross-validation and do a hypothesis test only for the holdout sample. 


```{r cars}
# read data
no_gender <- read.csv("http://dept.stat.lsa.umich.edu/~bbh/s485/data/emdstudy3-small-nogender.csv")
gender <- read.csv("http://dept.stat.lsa.umich.edu/~bbh/s485/data/emdstudy3-small.csv")

# renaming columns 
data_no_gender <- c("intel_theory", "attn_to", "ActPerc", "EstPerc") # this is for data w/o gender 
names(no_gender) <- data_no_gender

data_col <- c("intel_theory", "gender", "attn_to", "ActPerc", "EstPerc") # this is for data w/ gender 
names(gender) <- data_col

# changing all categorical variables in both datasets to factors
no_gender$attn_to <- as.factor(no_gender$attn_to)
gender$attn_to <- as.factor(gender$attn_to)

# fitted model w/o gender
mod0 <- lm(EstPerc - ActPerc ~ ., data = no_gender)
summary(mod0)
mean(mod0$residuals^2) 

par(mfrow=c(2,2)) # checking residuals: Normality, Homoscedasticity, Influential Points
plot(mod0)
points(fitted(mod0), resid(mod0), col = c("blue", "red"), pch = 16)
legend("topright", legend = c("Male", "Female"), col = c("blue", "red"), pch = 16)

# combine genders by independent variables 
ggplot(gender, aes(x = gender, y = intel_theory, fill = gender)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Comparing Intel Theory")

ggplot(gender, aes(x = gender, y = attn_to, fill = gender)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title="Comparing Attention")

ggplot(gender, aes(x = gender, y = ActPerc, fill = gender)) +
  geom_violin() +
  theme_minimal() +
  labs(title="Comparing Actual Percentile Score")

ggplot(gender, aes(x = gender, y = EstPerc, fill = gender)) +
  geom_violin() +
  theme_minimal() +
  labs(title="Comparing Estimated Percentile Score")

```


```{r models}
# model w/ gender 
mod1<- lm(EstPerc - ActPerc ~ ., data = gender)
summary(mod1)
plot(mod1)
mean(mod1$residuals^2) 

# compare the 2 models by checking the MSE & in-sample loss  
mse_mod0 <- mean(mod0$residuals^2)
mse_mod1 <- mean(mod1$residuals^2)
MSE_Diff <- mse_mod0 - mse_mod1
MSE_Diff

r_squared <- summary(mod1)$r.squared - summary(mod0)$r.squared 
r_squared

```


```{r tests}
# F test for both models 
anova(mod0, mod1)
```


# Additional calculations for paper version 2

```{r}
holdout_gender <- read.csv(file = "http://dept.stat.lsa.umich.edu/~bbh/s485/data/emdstudy3-all.csv", skip = 1, head =FALSE, sep=",")

names(holdout_gender) <- data_col

holdout_no_gender <- read.csv(file = "http://dept.stat.lsa.umich.edu/~bbh/s485/data/emdstudy3-all.csv", skip = 1, head =FALSE, sep=",")

holdout_no_gender <- holdout_no_gender[,-2]

names(holdout_no_gender) <- data_no_gender
```


```{r}
m1 <- lm(EstPerc - ActPerc ~ ., data = holdout_no_gender)
summary(m1)
plot(m1)
mean(m1$residuals^2)
```

```{r}
m2 <- lm(EstPerc - ActPerc ~ ., data = holdout_gender)
summary(m2)
plot(m2)
mean(m2$residuals^2)
```

```{r}
MSE_Diff = mean(m1$residuals^2) - mean(m2$residuals^2)
MSE_Diff
r_squared_diff = 0.04806 - 0.01207
r_squared_diff
```



```{r}
holdout_gender <- read.csv(file = "http://dept.stat.lsa.umich.edu/~bbh/s485/data/emdstudy3-all.csv", 
                           skip = 1, header = FALSE, sep = ",")

# column names
data_col <- c("intel_theory", "gender", "attn_to", "ActPerc", "EstPerc")  
colnames(holdout_gender) <- data_col

# w/o gender
holdout_no_gender <- holdout_gender[, !(colnames(holdout_gender) %in% "gender")]

# holdout data set only
holdout <- read.csv("http://dept.stat.lsa.umich.edu/~bbh/s485/data/emdstudy3-holdout.csv")

# Cross-validation
cv.lm <- function(data, formulae, nfolds = 5) {
  data <- na.omit(data)
  formulae <- sapply(formulae, as.formula)
  n <- nrow(data)
  fold.labels <- sample(rep(1:nfolds, length.out = n))
  mses <- matrix(NA, nrow = nfolds, ncol = length(formulae))
  colnames(mses) <- as.character(formulae)
  
  for (fold in 1:nfolds) {
    test.rows <- which(fold.labels == fold)
    train <- data[-test.rows, ]
    test <- data[test.rows, ]
    
    for (form in 1:length(formulae)) {
      current.model <- lm(formula = formulae[[form]], data = train)
      predictions <- predict(current.model, newdata = test)
      test.responses <- eval(formulae[[form]][[2]], envir = test)
      test.errors <- test.responses - predictions
      mses[fold, form] <- mean(test.errors^2)
    }
  }
  return(colMeans(mses))
}

# models
model_w_gender <- "EstPerc - ActPerc ~ gender + intel_theory + attn_to"
model_wo_gender <- "EstPerc - ActPerc ~ intel_theory + attn_to"

cv_results <- cv.lm(holdout_gender, c(model_w_gender, model_wo_gender))
print(cv_results)

# model fitting
m1 <- lm(EstPerc - ActPerc ~ ., data = holdout_no_gender)  # w/o gender
m2 <- lm(EstPerc - ActPerc ~ ., data = holdout_gender)  # w/ gender

# MSE
mse_w_gender <- mean((holdout$EstPerc - predict(m2, holdout_gender))^2)
mse_wo_gender <- mean((holdout$EstPerc - predict(m1, holdout_no_gender))^2)

mse_w_gender
mse_wo_gender

```


```{r}
# fitted models for holdout sample 
holdout_mod0 <- lm(EstPerc - ActPerc ~ intel_theory + attn_to, data = holdout_no_gender)
holdout_mod1 <- lm(EstPerc - ActPerc ~ gender + intel_theory + attn_to, data = holdout_gender)

anova(holdout_mod0, holdout_mod1)

```



